---
title: "STAT 184 Final Project: Exploratory Data Analysis"
output: html_notebook
author: "Jordan Williard"
date: "April 7th, 2020"
---

# Guiding Question (i.e., Research Question): Make a clear statement of a single guiding/research question that will drive your analysis.  It should be phrased as a question that you will try to investigate (i.e., it should have a question mark) 

How has the landscape of American political ideology changed throughout the past 100 years? Which states have experienced the most change? Are there any notable long term trends?

# Become acquainted with your data sources:
* Where did you find them?

    Data Source 1: Comparative Political Dataset (https://www.cpds-data.org/). I found this by searching and comparing various political datasets. 
  
    Data Source 2: US Population by Zip Code (https://data.world/lukewhyte/us-population-by-zip-code-2010-2016. I want to include demographic data in this analysis. This             dataset has a good cross section of years leading up to present day politics. 

* Who collected/maintains them?
    
    Data Source 1 is maintained by Prof. Dr. Klaus Armingeon and collaborators at the University of Zurich. 
    
    Data Source 2 was aggregated by a user on data.world. The data comes from 5-year American Surveys and the 2010 Decennial Census

* When & Why were they originally collected?
    
    The data in Data Source 1, the CPDS, has been collected since 1960 in many democratic countries. The goal is to better understand the political landscape of democratic       countries across the globe. I'll be focusing on the United States. 
    
    The data in Data Source 2 has been collected by the United States government for demographic purposes. These surveys allow the government to better understand who and        where people are living. 

* What does a case represent in each data source, and how many total cases are available?

    In Data Source 1, a case represents the political, social, and demographic information for one country, during one year. 
    
    In Data Source 2, a case represents the population of a zip code for years 2010-2016.

* What are some of the variables that you plan to use

    Data Source 1 has over 300 columns, so I will need to be selective with what I use. That being said, I am most interested in understanding changes in the American            political system as well as what opinions may be causing them. So, to start I am going to use: Total population, elderly population, political spectrums (varying           categorical degrees of left --> right), religion metrics, and protest information (could be indicative of unrest/revolution). This is just a start and I imagine that       I'll end up using many more.
    
    Simply put, I'm going to use all of the variables from Data Source 2. This data is extremely narrow, but will allow me to better understand demographic transitions         across regions of the United States. So, I will be using the Zip Code variable as well as all 7 years of population information. 


# Explore intuition related to the research question

### Loading Data
```{r}

library(DataComputing)

CPDS <- read.csv("CPDS-1960-2017-Update-2019.csv")
ZipPop <- read.csv("pop-by-zip-code.csv")

head(CPDS)
head(ZipPop)
```

* Create some informative plots and summary statistics 


CPDS: Simple summary statisitcs
```{r}
CPDS <-
  CPDS %>%
  select(year, country, realgdpgr, inflation, pop, elderly) %>%
  filter(country == "USA")

summary(CPDS)
CPDS$year[CPDS$inflation == min(CPDS$inflation)]
CPDS$year[CPDS$realgdpgr == min(CPDS$realgdpgr, na.rm = TRUE)]
CPDS$year[CPDS$elderly == max(CPDS$elderly, na.rm = TRUE)]
CPDS$year[CPDS$elderly == min(CPDS$elderly, na.rm = TRUE)]
```

ZipPop: Simple summary statisitcs
```{r}
summary(ZipPop)
mean(ZipPop$y.2010)
mean(ZipPop$y.2016)
```

```{r}
ZipPopMean <-
  ZipPop %>%
  mutate("mean2010" = mean(y.2010),
         "mean2011" = mean(y.2011),
         "mean2012" = mean(y.2012),
         "mean2013" = mean(y.2013),
         "mean2014" = mean(y.2014),
         "mean2015" = mean(y.2015),
         "mean2016" = mean(y.2016),
         ) %>%
  select(mean2010, mean2011, mean2012, mean2013, mean2014, mean2015, mean2016)

 avg_sizes <- ZipPopMean[!duplicated(ZipPopMean$mean2010),]
 
 avg_sizes
```


CPDS: Simple Graphs
```{r}
CPDS %>%
  ggplot(aes(x=year, y=elderly)) + 
  geom_line()
```

```{r}
CPDS %>%
  ggplot(aes(x=year, y=inflation)) + 
  geom_line()
```


```{r}
CPDS %>%
  ggplot(aes(x=year, y=realgdpgr)) + 
  geom_line()
```






* Describe preliminary observations and intuition about the research question

The following are some observations I've made by exploring this data:

    * The proportion of elderly in the United States has been steadily increasing. Age is an important determinant to political ideology, so I think that this will be important.
    
    * Different metrics that represent the economic situation in the United States have not been largely correlated over the years. A decrease in GDP and an increase in inflation is indicative of a situation impacting all segments of society. However, a decrease in GDP without an increase in inflation is indicative of a market       acting less favorable towards business owners/manufacturers. The disparity in these two metrics throughout the years, coupled with observations of political ideology   will tell me how different financial events impact politics. 
    
    * ZipPopMean tells us that the average number of residents reported since 2010 has not significantly gone up. This tells us that the Unites States population is largely stagnant (can be confirmed with other sources). That being said, it furthers my interest in changes for particular regions as well as the impacts that survey methodology may produce (immigrants not participating in a Census for example).






