---
title: "Final Project"
author: "Jordan Williard"
date: "4/26/2020"
output: html_document
---

### Workspace Initialization

```{r}
library(DataComputing)
library(tidyr)
```



### Introduction
America is on shaky political ground. The gap between left and right ideologies has been increasing to the point that the resulting tensions are almost to be expected. Fundamental differences between the ways that people view religion, abortion, immigration, and gun ownership, just to name a few, have created a polarized climate. As we approach the upcoming election cycle, it's valuable for campaigns and constituents alike to understand the ideological differences and shifts that are occuring in this nation. The following analysis is intended to observe these trends and answer the follow questions: What demographic trends exist in the United States and how do they relate to ideology within government?


### Data Access

Two data sets will be used for this analysis. The first is called the Comparative Political Data Set and contains information on demographic, political, and ideological characteristics for 36 countries from 1960 to 2017. In this data set, a case represents the political, social, and demographic information for one country, during one year. The second data set contains data collected 5 Year Surveys and Censi that has been aggregated by data.world. This data set can be used to observe population trends across America throughout the years. In this data set, a case represents the population for one zip code for years 2020-2016. 

```{r}
CPDS <- read.csv("CPDS-1960-2017-Update-2019.csv")
ZipPop <- read.csv("pop-by-zip-code.csv")

# head(CPDS) This is commented out simply to create a cleaner html document. Since the data has 319 columns, it takes up a lot of space when knit to html. 
head(ZipPop)
```


## Data Wrangling

### 1
Because I am only interested in American politics, I am going to begin by filtering the CPDS dataframe to only include cases pertaining to the US. There are also many columns that will not be useful for this analysis. The relevant fields are gathered with select(). Descriptions for these fields are as follows:

* 1. year: Year of data collection.

* 2. country: Country that data pertains to (Only USA after filtering).

* 3. gov_right1: The percentage of cabinet seats held by right wing politicians.

* 4. gov_cent1: The percentage of cabinet seats held by centrist politicians.

* 5. vturn: Voter turnout in election. 

* 6. pop: Total country population

* 7. elderly: Percentage of the population that is 65 years or older. 

* 8. unemp: Amount of civilians of working age who are unemployed. 

* 9. debt: Amount of debt accumulated in the year, expressed as percent of GDP.

* 10. realgdpgr: Percent growth of GDP as compared to the previous year.

An interesting observation from this is that left wing government composition is always zero in this data. Within America, this may not make sense, but on the world stage America does not have a left leaning arm of government. For this reason, I'm only gathering central and right ideology. 

```{r}
CPDS_US <-
  CPDS %>% 
  filter(country == "USA") %>%
  select(year, country, gov_right1, gov_cent1, vturn, pop, elderly, unemp, debt, realgdpgr)

head(CPDS_US)
```

### 2
By using machine learning to create a linear regression on this data, we observe several interesting relationships in the data. Here, elderly, unemployment rates, debt, and GDP data is being used to predict the prevalence of right wing ideologies in governemnt. We can see that the amount of elderly population as well as the amount of debt acrued in a single year are statistically significant. This tells us that an increase in the elderly population correlates to an increase in rightwing ideologies in government and that an increase in debt (as compared to GDP) also correlates to an increase of rightwing ideology within government. 

```{r}
linearMod <- lm(gov_right1 ~ elderly + unemp + debt + realgdpgr, data=CPDS_US)

summary(linearMod)

```

### 3 
When looking at demographic data, one of the first considerations is whether or not the total population of a region is increasing or decreasing. One way to do this is by determining if the average population by zip code changes throughout the years. By using the reduction function mean(), we can observe this trend and notice that the average population has remained fairly stagnant since 2010, indicating that the total population of the United States has also been fairly constant in this time period. Further research (https://www.worldometers.info/world-population/us-population/) confirms this observation: Between 2010 and 2016, the total population of the United States has only risen from 309M to 323M, a 4% increase.  

```{r}
ZipPopMean <-
  ZipPop %>%
  mutate("mean2010" = mean(y.2010),
         "mean2011" = mean(y.2011),
         "mean2012" = mean(y.2012),
         "mean2013" = mean(y.2013),
         "mean2014" = mean(y.2014),
         "mean2015" = mean(y.2015),
         "mean2016" = mean(y.2016),
         ) %>%
  select(mean2010, mean2011, mean2012, mean2013, mean2014, mean2015, mean2016)

 avg_sizes <- ZipPopMean[!duplicated(ZipPopMean$mean2010),]
 
 avg_sizes
```

### 4
The data in ZipPop is obviously not in a tidy format. In its initial state, a case represents a zip codes population across all years (2010-2016). Fortunately, a call to the wrangling function gather() can transform this into tidy data in which each case represents the population for a single zip code, in a single year. 

```{r}
ZipPopNarrow <-
  ZipPop %>%
  gather(Year, Population, -zip_code)
head(ZipPopNarrow)
nrow(ZipPopNarrow)
```

### 5
While the gather() function helped create tiny data, when the column headers for years were turned into keys they kept the leading "y." which will eliminate the use of any arithmetic pertaining to years. To resolve this, a regular expression will be used to remove any characters from the Year column that are not numeric. 

```{r}
ZipPopNarrow <-
  ZipPopNarrow %>%
  mutate(Year = gsub(pattern = "[^0-9-]", "", Year))
head(ZipPopNarrow)
```

## Data Visualizations

### 1
Based on the earlier significance found between debt and right wing ideologies, I wanted to get a better idea of how this has changed over time. The following bar graph shows the change in annual government debt with relation to year. Additionally, gov_right1 is added as an aesthetic so that we can see how the presense of ideologies in government relates to the accumulation of debt. The first interesting observation to note is that party lines shift roughly every 8 years. This makes sense based on the term limits that Presidents face along with the fact that most Presidents serve both of their terms. Furthermore, we can see a massive accumulation in debt in 2008. This isn't surprising based on the market crash that year, but it is interesting to see its relation to government spending. Remember, "debt" in this case is the percent of GDP that is borrowed. In this way, the graph tells us that the government has been borrowing a significant amount of money since 2008. Finally, with exception of Gerald Ford in the mid to late 70s, it's interesting to see that increased governmental borrowing is correlated with right wing governments for the entire span of this data. 

```{r}
CPDS_US %>%
  ggplot(aes(x=year,y=debt ,fill=gov_right1))+
  geom_bar(stat='identity',position='stack', width=1)

```

### 2
