---
title: "Final Project"
author: "Jordan Williard"
date: "4/26/2020"
output: html_document
---

### Workspace Initialization

```{r}
library(DataComputing)
library(tidyr)
```



### Introduction
America is on shaky political ground. The gap between left and right ideologies has been increasing to the point that the resulting tensions are almost to be expected. Fundamental differences between the ways that people view religion, abortion, immigration, and gun ownership, just to name a few, have created a polarized climate. As we approach the upcoming election cycle, it's valuable for campaigns and constituents alike to understand the ideological differences and shifts that are occuring in this nation. The following analysis is intended to observe these trends and answer the follow questions: What demographic trends exist in the United States and how do they relate to ideology within government? **GitHub Repo URL: https://github.com/jordanwilliard/STAT184_FinalProject **


### Data Access

Two data sets will be used for this analysis. The first is called the Comparative Political Data Set and contains information on demographic, political, and ideological characteristics for 36 countries from 1960 to 2017. In this data set, a case represents the political, social, and demographic information for one country, during one year. The second data set contains data collected 5 Year Surveys and Censi that has been aggregated by data.world. This data set can be used to observe population trends across America throughout the years. In this data set, a case represents the population for one zip code for years 2010-2016. 

```{r}
CPDS <- read.csv("CPDS-1960-2017-Update-2019.csv") # Loading data
ZipPop <- read.csv("pop-by-zip-code.csv")

# head(CPDS) This is commented out simply to create a cleaner html document. Since the data has 319 columns, it takes up a lot of space when knit to html. 
head(ZipPop)
```


## Data Wrangling

### 1
Because I am only interested in American politics, I am going to begin by filtering the CPDS dataframe to only include cases pertaining to the US. There are also many columns that will not be useful for this analysis. The relevant fields are gathered with select(). Descriptions for these fields are as follows:

* 1. year: Year of data collection.

* 2. country: Country that data pertains to (Only USA after filtering).

* 3. gov_right1: The percentage of cabinet seats held by right wing politicians.

* 4. gov_cent1: The percentage of cabinet seats held by centrist politicians.

* 5. vturn: Voter turnout in election. 

* 6. pop: Total country population

* 7. elderly: Percentage of the population that is 65 years or older. 

* 8. unemp: Amount of civilians of working age who are unemployed. 

* 9. debt: Amount of debt accumulated in the year, expressed as percent of GDP.

* 10. realgdpgr: Percent growth of GDP as compared to the previous year.

An interesting observation from this is that left wing government composition is always zero in this data. Within America, this may not make sense, but on the world stage America does not have a left leaning arm of government. For this reason, I'm only gathering central and right ideology. 

```{r}
CPDS_US <-
  CPDS %>% 
  filter(country == "USA") %>%                                                                  # Filter out cases not pertaining to USA
  select(year, country, gov_right1, gov_cent1, vturn, pop, elderly, unemp, debt, realgdpgr)     # Only gather fields that are relevant to research question

head(CPDS_US)
```

### 2
By using machine learning to create a linear regression on this data, we observe several interesting relationships in the data. Here, elderly, unemployment rates, debt, and GDP data is being used to predict the prevalence of right wing ideologies in government. We can see that the amount of elderly population as well as the amount of debt acrued in a single year are statistically significant. This tells us that an increase in the elderly population correlates to an increase in rightwing ideologies in government and that an increase in debt (as compared to GDP) also correlates to an increase of rightwing ideology within government. 

```{r}
linearMod <- lm(gov_right1 ~ elderly + unemp + debt + realgdpgr, data=CPDS_US)   # Create a linear regession model 

summary(linearMod)

```

### 3 
When looking at demographic data, one of the first considerations is whether or not the total population of a region is increasing or decreasing. One way to do this is by determining if the average population by zip code changes throughout the years. By using the reduction function mean(), we can observe this trend and notice that the average population has remained fairly stagnant since 2010, indicating that the total population of the United States has also been fairly constant in this time period. Further research (https://www.worldometers.info/world-population/us-population/) confirms this observation: Between 2010 and 2016, the total population of the United States has only risen from 309M to 323M, a 4% increase.  

```{r}
ZipPopMean <-
  ZipPop %>%
  mutate("mean2010" = mean(y.2010),                                   # Create fields that store the average zip code opulation by year
         "mean2011" = mean(y.2011),
         "mean2012" = mean(y.2012),
         "mean2013" = mean(y.2013),
         "mean2014" = mean(y.2014),
         "mean2015" = mean(y.2015),
         "mean2016" = mean(y.2016),
         ) %>%
  select(mean2010, mean2011, mean2012, mean2013, mean2014, mean2015, mean2016)

 avg_sizes <- ZipPopMean[!duplicated(ZipPopMean$mean2010),]          # Remove any duplicate data
 
 avg_sizes
```

### 4
The data in ZipPop is obviously not in a tidy format. In its initial state, a case represents a zip codes population across all years (2010-2016). Fortunately, a call to the wrangling function gather() can transform this into tidy data in which each case represents the population for a single zip code, in a single year. 

```{r}
ZipPopNarrow <-
  ZipPop %>%
  gather(Year, Population, -zip_code)   # Gather the cases to create tidy data in which each case represents zip code population by year
head(ZipPopNarrow)
nrow(ZipPopNarrow)
```

### 5
While the gather() function helped create tiny data, when the column headers for years were turned into keys they kept the leading "y." which will eliminate the use of any arithmetic pertaining to years. To resolve this, a regular expression will be used to remove any characters from the Year column that are not numeric. 

```{r}
ZipPopNarrow <-
  ZipPopNarrow %>%
  mutate(Year = gsub(pattern = "[^0-9-]", "", Year))   # Use a regex to remove any characters from the Year field that are not numeric
head(ZipPopNarrow)
```

## Data Visualizations

### 1
Based on the earlier significance found between debt and right wing ideologies, I wanted to get a better idea of how this has changed over time. The following bar graph shows the change in annual government debt with relation to year. Additionally, gov_right1 is added as an aesthetic so that we can see how the presense of ideologies in government relates to the accumulation of debt. The first interesting observation to note is that party lines shift roughly every 8 years. This makes sense based on the term limits that Presidents face along with the fact that most Presidents serve both of their terms. Furthermore, we can see a massive accumulation in debt in 2008. This isn't surprising based on the market crash that year, but it is interesting to see its relation to government spending. Remember, "debt" in this case is the percent of GDP that is borrowed. In this way, the graph tells us that the government has been borrowing a significant amount of money since 2008. Finally, with exception of Gerald Ford in the mid to late 70s, it's interesting to see that increased governmental borrowing is correlated with right wing governments for the entire span of this data. 

```{r}
CPDS_US %>%
  ggplot(aes(x=year,y=debt ,fill=gov_right1))+         # Aesthetics for the variables of interest
  geom_bar(stat='identity',position='stack', width=1) + 
  geom_line()

```

### 2
Finally, I am interested in seeing the specifics of how economic and demographic factors are related to which party holds power in office. In order to do this, an additional column is created to store whether or not the government is controlled (>= 50%) by right wing ideology. Because right wing ideology pertains to the Republican party in the US, these instances hold the value "TRUE" for the field Republican. With a classification system in this data, we can now generate a decision tree using several important characterstics to predict whether or not Republicans will hold power. For this analysis, voter turnout, percent of elderly, percent of unemployment, and debt are being used in an effort to capture demographic and economic factors. As the decision tree is generated and displayed, the following observations are made: At the highest levels of elderly population, the Republican party holds office. However, for every other cut of elderly population, the party varies. At the lowest levels of elderly populations, debt is considered. Here we also see that higher levels of borrowing within a government relate to the Republican party. Interestingly enough, voter turnout and unemployment do not have an impact on this tree which further confirms the statistical significance found above.  

```{r}
CPDS_US_Party <-
  CPDS_US %>%
  mutate(Republican = if_else(gov_right1 >= 50, TRUE, FALSE)) %>%     # Create an additional field to determine whether or not Republicans hold power in that year
  select(Republican, vturn, elderly, unemp, debt, year) %>%
  arrange(year)

head(CPDS_US_Party,20)
```

```{r}
library(rpart)                                               # Load library to create a decision tree

fit <- rpart(Republican ~ vturn + elderly + unemp + debt,    # Fit the decision tree to the new data frame
   method="class", data=CPDS_US_Party)

plot(fit, uniform=TRUE,                                      # Visualize the decision tree to understand its rules
   main="Classification Tree for Republican Party\n")
text(fit, use.n=TRUE, all=TRUE, cex=0.7)

```

## Conclusion

This analysis resulted in a more in depth understanding of how demographic, economic, and ideological factors play a part in America's everchanging political landscape. We can see that above all, the amount of money that a government borrows in a year combined with the amount of elderly people in the population have notable effects on who is in office. One of my goals at the onset of this project was to gain a better understanding of who (or which party) has the best chance of winning the 2020 election. While it's not always easy to predict political events, this project has shown me that the Republican party (Donald Trump) will likely win the next election for a couple of reasons. First and foremost, COVID has had significant financial implications. Millions have become unemployed and the Treasury released a massive stimulus package (AKA borrowing money). Additionally, America's elderly white population continues to increase in number. These factors alone lead me to the conclusion that the Republican party has a better shot in 2020. All in all, this project gave me the opportunity to apply the knowledge I've gained throughout STAT 184 to a topic that I am interested in. Data that has been gathered from the web is not always in a tidy, or even "good", format and leveraging the skills gained in this class I have been able to wrangle, analyze, and train classifiers on this data to learn more about American politics and society.  
